---
title: "Global Energy Transition and the Emissions Footprint"
subtitle: "A Visual Analysis of How, Where, and Why the World Decarbonises Unevenly"
author: "Group 7: Hrusheekesh Sawarkar, Kseniia Holovchenko, Melisa Cihan, Danish Ali Abdul Kareem Shaikh,  Amaan Aslam Shaikh"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    fig-width: 9
    fig-height: 5.5
    theme: cosmo
    embed-resources: true
    fig-cap-location: bottom
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    fig-width: 9
    fig-height: 5.5
    fig-cap-location: bottom
execute:
  warning: false
  message: false
  cache: true
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(scales)
library(ggrepel)
library(patchwork)
library(countrycode)
library(sf)
library(rnaturalearth)
library(gifski)
library(gganimate)
library(magick)

theme_report <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 14),
      plot.subtitle    = element_text(colour = "grey30"),
      plot.caption     = element_text(colour = "grey55", size = 8),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
}
theme_set(theme_report())

# Colour-blind-checked palette (Okabe-Ito-derived where possible). Solar and the
# two "renewables" categories are deliberately separated in hue and lightness so
# that adjacent stacked bands remain distinguishable under deuteranopia.
energy_pal <- c(
  Coal              = "#3d3d3d",
  Oil               = "#8c5a33",
  Gas               = "#e08214",
  Nuclear           = "#6a51a3",
  Hydro             = "#2166ac",
  Wind              = "#56b4e9",
  Solar             = "#f0c000",
  `Other renewables`= "#1b7837",
  Renewables        = "#1b7837"
)

# Fixed continent colours (Set2-derived), keyed by name so the mapping is
# independent of factor-level order. Both continent figures (the area chart and
# the emitter ranking) draw from this vector, so a continent keeps the same
# colour across figures even though each figure orders its own legend differently.
continent_pal <- c(
  Asia     = "#A6D854",
  Americas = "#E78AC3",
  Europe   = "#8DA0CB",
  Africa   = "#FC8D62",
  Oceania  = "#66C2A5"
)

owid_caption <- "Source: Our World in Data, energy and CO2 datasets"

lab_co2_pc <- expression(CO[2]~"per capita (tonnes)")
lab_co2_mt <- expression(CO[2]~"(million tonnes)")
lab_co2_gt <- expression(CO[2]~"(gigatonnes)")
```

# Introduction {#sec-intro}

The transition from fossil fuels to low-carbon energy is the defining infrastructural undertaking of the present century. It is, however, frequently discussed as though it were a single global event, when in practice it proceeds at markedly different rates across countries, and even identical historical shocks elicit divergent national responses. The objective of this report is to render that unevenness analytically visible and to identify the structural factors that explain it.

The analysis draws on two open panel datasets published by Our World in Data (OWID): the [Energy dataset](https://github.com/owid/energy-data) and the [CO2 and Greenhouse-Gas dataset](https://github.com/owid/co2-data). Because the two sources share `iso_code` and `year` as keys, energy structure can be joined directly to the emissions that arise from it, allowing hypotheses about the transition to be examined against a consistent global record. To guarantee reproducibility, both files are downloaded once, cached locally, and read from disk for every render; the access date is recorded below so that the figures and the numbers quoted in the prose remain mutually consistent regardless of subsequent upstream updates.

```{r load-data}
#| label: load-data

# --- Reproducible data acquisition -----------------------------------------
# Download each CSV once into a local cache, then always read from disk. This
# pins the analysis to a fixed snapshot: re-rendering next month cannot silently
# change the figures or invalidate the numbers quoted in the text. Delete the
# cache directory to refresh against the latest upstream release.

data_dir <- "data"
dir.create(data_dir, showWarnings = FALSE)

energy_path <- file.path(data_dir, "owid-energy-data.csv")
co2_path    <- file.path(data_dir, "owid-co2-data.csv")

energy_url <- "https://github.com/owid/energy-data/raw/master/owid-energy-data.csv"
co2_url    <- "https://github.com/owid/co2-data/raw/master/owid-co2-data.csv"

if (!file.exists(energy_path)) download.file(energy_url, energy_path, mode = "wb")
if (!file.exists(co2_path))    download.file(co2_url,    co2_path,    mode = "wb")

data_accessed <- format(file.info(co2_path)$mtime, "%Y-%m-%d")

energy_raw <- read_csv(energy_path, show_col_types = FALSE)
co2_raw    <- read_csv(co2_path,    show_col_types = FALSE)

# OWID interleaves genuine countries with regional and income aggregates that
# carry a blank or "OWID_"-prefixed iso_code. These are removed for all
# per-country analysis to prevent double counting, and reintroduced by name
# (e.g. country == "World") only where an aggregate total is required.
is_real_country <- function(code) {
  !is.na(code) & str_detect(code, "^[A-Z]{3}$") & !str_starts(code, "OWID_")
}

energy <- energy_raw |> filter(is_real_country(iso_code))
co2     <- co2_raw    |> filter(is_real_country(iso_code))
```

The data used here were accessed on `r data_accessed`. All figures and quoted statistics derive from that snapshot.

# Exploratory Data Analysis {#sec-eda}

Before any substantive argument is developed, it is necessary to establish how the principal variables are distributed and how they relate to one another. The exploratory views below characterise the scope, coverage, distributional shape, correlation structure, and low-dimensional geometry of the data and, in doing so, motivate the analytical choices made in the remainder of the report.

A preliminary point about coverage governs the choice of reference year. The GDP series originates in the Maddison Project Database, whose most recent observations are both sparse and provisional, so the latest year for which energy, emissions, and GDP are jointly well populated is earlier than the final year of the emissions series. The reference year for every cross-sectional figure is therefore selected automatically as the most recent year in which a sufficient number of countries report all four core indicators, rather than fixed by hand, which prevents the decoupling analysis from being computed on an unrepresentatively thin sample.

```{r eda-refyear}
#| label: eda-refyear
#| include: false

# Choose the most recent year in which a reasonable number of countries
# (population > 1e6) jointly report GDP, CO2 per capita, energy per capita and
# renewable share. Energy variables live in the energy table, emissions/GDP in
# the co2 table, so each is sourced from the frame that actually carries it.
core <- co2 |>
  filter(population > 1e6) |>
  transmute(iso_code, year,
            gdp_pc = gdp / population,
            co2_pc = co2_per_capita) |>
  inner_join(
    energy |> select(iso_code, year,
                     ene_pc = energy_per_capita,
                     ren    = renewables_share_energy),
    by = c("iso_code", "year")
  )

coverage <- core |>
  filter(if_all(c(gdp_pc, co2_pc, ene_pc, ren), ~ !is.na(.x))) |>
  count(year)

# Most recent year clearing a coverage floor; relax the floor if no year does,
# and fall back to the best-covered year so eda_year is never empty.
pick_year <- function(tbl, floor) {
  tbl |> filter(n >= floor) |> slice_max(year, n = 1) |> pull(year)
}
ref_year <- pick_year(coverage, 100)
if (length(ref_year) == 0) ref_year <- pick_year(coverage, 60)
if (length(ref_year) == 0) ref_year <- coverage |> slice_max(n, n = 1) |> pull(year)

eda_year <- ref_year[1]   # reused by every cross-sectional figure below
stopifnot(length(eda_year) == 1, !is.na(eda_year))

ref_n <- coverage |> filter(year == eda_year) |> pull(n)
```

The most recent year meeting the joint-coverage criterion is **`r eda_year`** (with `r ref_n` countries reporting all four indicators), and it is used as the common reference year for all cross-sectional figures so that they remain directly comparable.

## Scope and shape of the data {#sec-scope}

The combined panel is first described at a high level: how far back the records extend, how many countries and observations they contain, and how completely each core variable is reported. These figures frame every claim that follows and expose where coverage is thin enough to qualify an interpretation.

```{r eda-scope}
#| label: eda-scope
#| include: false

# Panel dimensions and span, computed from the cleaned per-country frames.
energy_span <- range(energy$year, na.rm = TRUE)
co2_span    <- range(co2$year,    na.rm = TRUE)
n_countries <- n_distinct(c(energy$iso_code, co2$iso_code))
n_energy_rows <- nrow(energy)
n_co2_rows    <- nrow(co2)

# Year in which wind and solar first reach 1% of global primary energy, which
# justifies restricting the renewables-era figures to the period from 2000.
ws_year <- energy_raw |>
  filter(country == "World", !is.na(wind_consumption), !is.na(solar_consumption),
         !is.na(primary_energy_consumption)) |>
  mutate(ws_share = 100 * (wind_consumption + solar_consumption) /
                    primary_energy_consumption) |>
  filter(ws_share >= 1) |>
  slice_min(year, n = 1) |>
  pull(year)
```

The cleaned per-country panel spans **`r energy_span[1]`–`r energy_span[2]`** for energy and **`r co2_span[1]`–`r co2_span[2]`** for emissions, covering roughly **`r n_countries`** countries. After removing OWID's regional and income aggregates, the energy frame holds about **`r format(n_energy_rows, big.mark = ",")`** country–year observations and the emissions frame about **`r format(n_co2_rows, big.mark = ",")`**. Although the emissions record reaches back to the nineteenth century, the cross-country figures in this report are confined to the well-covered modern era, and the figures that depend on wind and solar generation begin at **`r ws_year`**, the first year in which those two technologies jointly supplied at least one percent of global primary energy. The structure of the two cleaned panels is summarised in @tbl-panel.

```{r eda-panel-table}
#| label: tbl-panel
#| tbl-cap: "Structure of the two cleaned per-country panels. The temporal range gives the first and last reporting year; observations are country–year rows after removal of OWID's regional and income aggregates; mean observations per country is the row count divided by the number of reporting countries; and mean completeness is the average non-missing rate across that panel's core indicators, indicating how much of the nominal grid is actually populated."

panel_overview <- tibble(
  Panel = c("Energy", "Emissions (CO2)"),
  # Years are stored as character so kable() does not insert thousands
  # separators (e.g. "2,025") via big.mark.
  `First year` = as.character(c(energy_span[1], co2_span[1])),
  `Last year`  = as.character(c(energy_span[2], co2_span[2])),
  Countries    = c(n_distinct(energy$iso_code), n_distinct(co2$iso_code)),
  Observations = c(n_energy_rows, n_co2_rows),
  `Mean obs. per country` = c(
    n_energy_rows / n_distinct(energy$iso_code),
    n_co2_rows    / n_distinct(co2$iso_code)),
  `Mean completeness (%)` = c(
    100 * mean(!is.na(as.matrix(energy[c("energy_per_capita",
                                         "primary_energy_consumption",
                                         "renewables_share_energy")]))),
    100 * mean(!is.na(as.matrix(co2[c("co2", "co2_per_capita",
                                      "co2_per_unit_energy", "gdp")]))))
) |>
  mutate(across(c(`Mean obs. per country`, `Mean completeness (%)`),
                ~ round(.x, 1)))

knitr::kable(panel_overview, format.args = list(big.mark = ","))
```

The contrast in @tbl-panel is informative in itself. The emissions panel runs far longer and holds more than twice as many observations, because OWID extends the CO2 record back into the eighteenth and nineteenth centuries for the early industrialisers, whereas the detailed energy-mix series begin only in the twentieth century. Mean completeness is below one half in both panels, confirming that the nominal country–year grid is sparsely populated and that the modern, well-covered window adopted throughout the report is a necessary restriction rather than a stylistic choice.

That sparsity has a temporal structure which a single completeness figure cannot convey, so the next figure traces, year by year, how many countries report each of four core indicators. It shows directly when the panel becomes dense enough to support cross-country comparison and why the reference year cannot simply be the final year of the record.

```{r eda-coverage}
#| label: fig-coverage
#| fig-cap: !expr 'paste0("Number of countries reporting each core indicator, 1900 to present. Emissions coverage widens gradually over the twentieth century; the energy series begin abruptly in 1965; the detailed renewables share is available for only about eighty countries in any year; and the GDP series (Maddison Project) terminates several years before the emissions and energy records do. The sharp fall in the energy line at the final year reflects incomplete reporting in the latest release rather than a genuine loss of coverage. The dashed line marks the automatically selected reference year, ", eda_year, ".")'
#| fig-height: 4

cov_long <- bind_rows(
  co2 |> filter(!is.na(co2_per_capita)) |> count(year) |>
    mutate(indicator = "CO2 per capita"),
  co2 |> filter(!is.na(gdp)) |> count(year) |>
    mutate(indicator = "GDP"),
  energy |> filter(!is.na(energy_per_capita)) |> count(year) |>
    mutate(indicator = "Energy per capita"),
  energy |> filter(!is.na(renewables_share_energy)) |> count(year) |>
    mutate(indicator = "Renewables share")
) |>
  filter(year >= 1900) |>
  mutate(indicator = factor(indicator,
                            levels = c("CO2 per capita", "GDP",
                                       "Energy per capita", "Renewables share")))

# Colour-blind-checked four-line palette; hues drawn from the report's existing
# scheme so the same indicator families keep familiar colours.
cov_pal <- c(`CO2 per capita`    = "#B2182B",
             GDP                 = "#c46210",
             `Energy per capita` = "#2166ac",
             `Renewables share`  = "#1b7837")

ggplot(cov_long, aes(year, n, colour = indicator)) +
  geom_vline(xintercept = eda_year, linetype = "dashed",
             colour = "grey55", linewidth = 0.4) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = cov_pal, name = NULL) +
  scale_x_continuous(breaks = seq(1900, 2020, 20)) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Reporting coverage is uneven in time as well as across countries",
       subtitle = "Countries reporting each indicator, by year",
       x = NULL, y = "Countries reporting", caption = owid_caption)
```

The figure explains three decisions made earlier. First, the energy indicators are entirely absent before 1965 and reach broad coverage only around 1980, which is why every energy-mix comparison in the report is confined to the modern era. Second, the renewables-share series never rises much above eighty countries, because the detailed fuel-mix data derive from the Energy Institute's *Statistical Review*, which covers only the larger energy economies; every analysis that conditions on renewable share therefore works with a smaller and somewhat wealthier sample than the emissions record suggests. Third, the GDP curve stops entirely at the right-hand edge: the Maddison series has simply not yet been compiled for the latest years. The reference year `r eda_year` is the last year before that termination, which is precisely what the automatic selection rule in @sec-eda was designed to achieve; fixing the reference year at the end of the emissions series instead would silently discard every economy from the GDP-dependent figures.

The next table moves from the structure of the panels to the seven indicators used throughout the report, summarising each for the chosen reference year by its centre, spread, and range alongside the number of countries reporting it. The gap between the mean and the median in several rows previews the right-skew documented in @sec-dist, and the coverage column shows which indicators are most affected by missingness.

```{r eda-summary-table}
#| label: tbl-summary
#| tbl-cap: !expr 'paste0("Descriptive statistics for the core indicators across countries with populations above one million, ", eda_year, ". ''n'' is the number of countries reporting the indicator; IQR is the interquartile range. Wide mean-versus-median gaps indicate right-skew.")'

summary_df <- co2 |>
  filter(year == eda_year, population > 1e6) |>
  transmute(iso_code,
            `GDP per capita (int-USD)` = gdp / population,
            `CO2 per capita (t)`       = co2_per_capita,
            `CO2 per GDP (kg/USD)`     = co2_per_gdp,
            `Carbon intensity (kg/kWh)`= co2_per_unit_energy) |>
  left_join(
    energy |> filter(year == eda_year) |>
      select(iso_code,
             `Energy per capita (kWh)` = energy_per_capita,
             `Renewable share (%)`     = renewables_share_energy,
             `Low-carbon share (%)`    = low_carbon_share_energy),
    by = "iso_code"
  ) |>
  select(-iso_code)

summary_tbl <- summary_df |>
  pivot_longer(everything(), names_to = "Indicator", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(Indicator) |>
  summarise(n      = n(),
            Mean    = mean(value),
            Median  = median(value),
            SD      = sd(value),
            IQR     = IQR(value),
            Min     = min(value),
            Max     = max(value),
            .groups = "drop") |>
  mutate(across(c(Mean, Median, SD, IQR, Min, Max),
                ~ signif(.x, 3)))

knitr::kable(summary_tbl, format.args = list(big.mark = ","))
```

A few features stand out even before the dedicated figures. Per-capita GDP, per-capita CO2, and energy use each span more than two orders of magnitude across countries, so their means sit well above their medians; the renewable and low-carbon shares, being bounded between zero and one hundred, are less extreme but still markedly uneven; and carbon intensity is comparatively compact, reflecting the shared physics of combustion. These shapes are the reason the report relies on medians, logarithmic axes, and distribution plots rather than on means alone.

## Distribution of the principal indicators {#sec-dist}

Summary means are unreliable descriptors when a variable is heavily skewed. The four quantities on which the report most depends are therefore examined directly for the reference year.

```{r eda-distributions}
#| label: fig-distributions
#| fig-cap: !expr 'paste0("Cross-country distributions for ", eda_year, ", restricted to countries with populations above one million. GDP per capita, CO2 per capita, and renewable share are all strongly right-skewed, with a small number of extreme economies stretching the upper tail.")'
#| fig-height: 5.5

dist_df <- co2 |>
  filter(year == eda_year, population > 1e6) |>
  transmute(iso_code,
            `GDP per capita (US$)` = gdp / population,
            `CO2 per capita (t)`   = co2_per_capita) |>
  left_join(
    energy |> filter(year == eda_year) |>
      select(iso_code,
             `Energy per capita (kWh)` = energy_per_capita,
             `Renewables share (%)`    = renewables_share_energy),
    by = "iso_code"
  ) |>
  pivot_longer(-iso_code, names_to = "variable", values_to = "value") |>
  filter(!is.na(value), value > 0)

ggplot(dist_df, aes(value)) +
  geom_histogram(bins = 30, fill = "#2166ac", colour = "white", alpha = 0.85) +
  facet_wrap(~variable, scales = "free", ncol = 2) +
  scale_x_continuous(labels = label_number(big.mark = ",")) +
  labs(title = "The principal indicators are right-skewed",
       subtitle = paste0("Distribution across countries above one million people, ",
                          eda_year),
       x = NULL, y = "Number of countries", caption = owid_caption)
```

```{r eda-dist-stats}
#| label: eda-dist-stats
#| include: false

dist_stats <- co2 |>
  filter(year == eda_year, population > 1e6) |>
  summarise(med_co2 = median(co2_per_capita, na.rm = TRUE),
            max_co2 = max(co2_per_capita, na.rm = TRUE))

ren_stats <- energy |>
  filter(year == eda_year) |>
  summarise(med_ren = median(renewables_share_energy, na.rm = TRUE))
```

Three of the four distributions exhibit a pronounced right tail. The median country emits about `r round(dist_stats$med_co2, 1)` tonnes of CO2 per capita, whereas several petro-states approach `r round(dist_stats$max_co2, 0)`; the median renewable share is near `r round(ren_stats$med_ren, 0)` percent, with only a limited cluster of leaders above fifty. This skewness is itself informative, since it indicates that the typical country differs substantially from the high-profile extremes, and it justifies the logarithmic axes adopted in @sec-decoupling, without which the numerous small economies would be compressed into an uninformative margin.

The pooled histograms conceal where in the world the tail countries sit, so the same emissions variable is next broken out by continent. This adds a first geographical dimension to the distributional story and anticipates the spatial analysis of @sec-geo.

```{r eda-boxplot}
#| label: fig-box-continent
#| fig-cap: !expr 'paste0("CO2 emissions per capita by continent, ", eda_year, ", for countries with populations above one million, on a logarithmic scale. Each point is a country; boxes give the median and interquartile range, and continents are ordered by their median. Oceania contains only a handful of countries above the population threshold, so its box rests on very few observations. The vertical spread within continents rivals the differences between them, and the most extreme emitters are the Gulf oil economies within Asia rather than any continent as a whole.")'
#| fig-height: 5

box_df <- co2 |>
  filter(year == eda_year, population > 1e6, !is.na(co2_per_capita),
         co2_per_capita > 0) |>
  mutate(continent = countrycode(iso_code, "iso3c", "continent")) |>
  filter(!is.na(continent)) |>
  mutate(continent = fct_reorder(continent, co2_per_capita,
                                 .fun = median, na.rm = TRUE))

box_label <- box_df |>
  group_by(continent) |>
  slice_max(co2_per_capita, n = 1) |>
  ungroup()

ggplot(box_df, aes(continent, co2_per_capita)) +
  geom_boxplot(aes(fill = continent), alpha = 0.45, colour = "grey35",
               outlier.shape = NA, width = 0.55) +
  geom_jitter(aes(colour = continent), width = 0.14, size = 1.6, alpha = 0.8) +
  geom_text_repel(data = box_label, aes(label = country),
                  size = 2.7, colour = "grey35", direction = "y",
                  nudge_x = 0.38, seed = 7) +
  scale_y_log10(labels = label_number()) +
  scale_fill_manual(values = continent_pal, guide = "none") +
  scale_colour_manual(values = continent_pal, guide = "none") +
  labs(title = "Within-continent spread rivals between-continent differences",
       subtitle = paste0("CO2 per capita by continent (log scale), ", eda_year,
                          "; the highest emitter per continent is labelled"),
       x = NULL, y = lab_co2_pc, caption = owid_caption)
```

The ordering of medians runs from Africa, where the median country emits well under one tonne per person, through the Americas and Asia to Europe and Oceania. Yet the within-continent spread is the more consequential feature: Asia alone spans the entire global range, containing both some of the lowest per-capita emitters and, in the Gulf states, the very highest. Continental membership is therefore a poor predictor of an individual country's footprint, which reinforces the report's emphasis on country-level structure over regional generalisation.

## Correlation structure of the core indicators {#sec-corr}

The correlation matrix below summarises the pairwise relationships among the report's central indicators. It serves both as a diagnostic, identifying which relationships warrant a dedicated figure, and as an early indication of the report's organising tension.

```{r eda-corr}
#| label: fig-corr
#| fig-cap: !expr 'paste0("Pairwise Pearson correlations across countries for ", eda_year, ", computed on complete cases. Warm cells denote positive associations and cool cells negative ones. Because the matrix is symmetric, only the lower triangle is shown; the diagonal of ones is omitted. The relationships that later receive dedicated figures are those with the strongest coefficients.")'
#| fig-height: 6

corr_df <- energy |>
  filter(year == eda_year) |>
  select(iso_code,
         `Renewable share`  = renewables_share_energy,
         `Low-carbon share` = low_carbon_share_energy,
         `Energy/capita`    = energy_per_capita,
         `Energy/GDP`       = energy_per_gdp) |>
  inner_join(
    co2 |> filter(year == eda_year) |>
      transmute(iso_code,
                `GDP/capita`       = gdp / population,
                `CO2/capita`       = co2_per_capita,
                `CO2/GDP`          = co2_per_gdp,
                `Carbon intensity` = co2_per_unit_energy),
    by = "iso_code"
  ) |>
  select(-iso_code) |>
  drop_na()

cmat <- cor(corr_df, use = "complete.obs")

# Coefficient that anchors the report's central question, quoted inline below.
r_gdp_ren <- cmat["GDP/capita", "Renewable share"]
# Spearman as a robustness check against the right-skew and hydro outliers.
rho_gdp_ren <- cor(corr_df$`GDP/capita`, corr_df$`Renewable share`,
                   method = "spearman")

# The matrix is symmetric, so only the lower triangle (excluding the diagonal
# of ones) is plotted; the dropped rows carry no additional information.
corr_long <- as.data.frame(cmat) |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  mutate(across(c(var1, var2), ~factor(.x, levels = rownames(cmat)))) |>
  filter(as.integer(var2) > as.integer(var1)) |>
  mutate(var1 = droplevels(var1), var2 = droplevels(var2))

ggplot(corr_long, aes(var1, var2, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3,
            colour = ifelse(abs(corr_long$r) > 0.6, "white", "grey20")) +
  scale_fill_gradient2(low = "#2166AC", mid = "grey95", high = "#B2182B",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  coord_fixed() +
  labs(title = "Cross-country correlation of energy and emissions indicators",
       subtitle = paste0("Pairwise Pearson correlations, ", eda_year),
       x = NULL, y = NULL, caption = owid_caption) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        panel.grid = element_blank())
```

Three features of the matrix are analytically consequential. First, wealthier countries consume substantially more energy per capita and, correspondingly, emit more per capita, reflecting the historical coupling of prosperity to fossil-fuelled industrial activity. Second, and more encouragingly, a higher renewable share is associated with markedly lower carbon intensity, indicating that altering the composition of energy supply genuinely reduces the emissions embodied in each unit of energy consumed. Third, and most importantly for what follows, the correlation between GDP per capita and renewable share is close to zero (Pearson *r* = `r sprintf("%.2f", r_gdp_ren)`, Spearman *ρ* = `r sprintf("%.2f", rho_gdp_ren)`, the two agreeing despite the skew). National wealth, in other words, does not determine how clean a country's energy system is. This near-absence of association is the empirical seed of the report's central question, addressed directly in @sec-decoupling: if income alone does not govern energy cleanliness, what does?

Because a single coefficient can conceal structure — a strong nonlinear relationship, or two offsetting clusters, can each produce a correlation near zero — the decisive cell of the matrix is examined directly as a scatterplot before any conclusion is drawn from it.

```{r eda-gdp-ren}
#| label: fig-gdp-ren
#| fig-cap: !expr 'paste0("GDP per capita (logarithmic) against the renewable share of primary energy, ", eda_year, ", for countries with populations above one million, coloured by continent. The fitted line is essentially flat and the cloud shows no hidden nonlinear pattern: rich and poor countries alike appear at every level of renewable share. Norway and Qatar sit at almost the same income yet at opposite extremes of the vertical axis.")'
#| fig-height: 5.5

gdp_ren <- co2 |>
  filter(year == eda_year, population > 1e6, !is.na(gdp)) |>
  transmute(iso_code, country, gdp_pc = gdp / population) |>
  filter(gdp_pc > 0) |>
  inner_join(
    energy |> filter(year == eda_year) |>
      select(iso_code, ren = renewables_share_energy),
    by = "iso_code"
  ) |>
  filter(!is.na(ren)) |>
  mutate(continent = countrycode(iso_code, "iso3c", "continent")) |>
  filter(!is.na(continent))

gdp_ren_label <- c("Norway", "Sweden", "Qatar", "United States", "China",
                   "India", "Germany", "Brazil", "Bangladesh")

ggplot(gdp_ren, aes(gdp_pc, ren)) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, colour = "grey40",
              fill = "grey85", linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(colour = continent), size = 2.2, alpha = 0.8) +
  geom_text_repel(data = filter(gdp_ren, country %in% gdp_ren_label),
                  aes(label = country), size = 2.9, colour = "grey30",
                  max.overlaps = Inf, box.padding = 0.35, seed = 7) +
  scale_x_log10(labels = dollar_format(accuracy = 1)) +
  scale_colour_manual(values = continent_pal, name = NULL) +
  labs(title = "Wealth does not buy a clean energy system",
       subtitle = paste0("GDP per capita (log) against renewable share, ", eda_year,
                          "; dashed line is the fitted linear trend"),
       x = "GDP per capita (logarithmic scale)",
       y = "Renewables share of primary energy (%)",
       caption = owid_caption)
```

The scatterplot confirms that the near-zero coefficient is genuine rather than an artefact of averaging. The fitted line is flat across two orders of magnitude of income, and no curvature or clustering emerges that a linear coefficient might have missed. The most instructive contrasts are vertical rather than horizontal: Norway and Qatar command almost identical GDP per capita yet occupy opposite ends of the renewable axis, while in the middle of the income range Brazil's hydroelectric endowment lifts its renewable share to nearly one half, far above similarly placed fossil-dependent economies. Whatever separates clean from carbon-intensive energy systems, it is visible at every income level, and it is not wealth itself.

## A low-dimensional view: principal component analysis {#sec-pca}

The correlation matrix reports pairwise associations, but it cannot show whether the indicators jointly reduce to a small number of independent dimensions. Principal component analysis (PCA) answers that question directly. Six indicators are used: GDP per capita, CO2 per capita, and energy per capita, each log-transformed on account of the right-skew documented in @sec-dist, together with the renewable share, the low-carbon share, and carbon intensity, which are bounded or compact and enter untransformed. All six are standardised before decomposition so that no single indicator dominates by virtue of its units.

```{r eda-pca}
#| label: eda-pca
#| include: false

pca_df <- co2 |>
  filter(year == eda_year, population > 1e6) |>
  transmute(iso_code, country,
            gdp_pc    = gdp / population,
            co2_pc    = co2_per_capita,
            intensity = co2_per_unit_energy) |>
  inner_join(
    energy |> filter(year == eda_year) |>
      select(iso_code,
             ene_pc = energy_per_capita,
             ren    = renewables_share_energy,
             lowc   = low_carbon_share_energy),
    by = "iso_code"
  ) |>
  # Complete cases only; the logged variables must additionally be positive.
  filter(if_all(c(gdp_pc, co2_pc, ene_pc), ~ !is.na(.x) & .x > 0),
         !is.na(ren), !is.na(lowc), !is.na(intensity)) |>
  mutate(continent = countrycode(iso_code, "iso3c", "continent")) |>
  filter(!is.na(continent))

pca_mat <- pca_df |>
  transmute(`GDP/capita (log)`    = log10(gdp_pc),
            `CO2/capita (log)`    = log10(co2_pc),
            `Energy/capita (log)` = log10(ene_pc),
            `Renewable share`     = ren,
            `Low-carbon share`    = lowc,
            `Carbon intensity`    = intensity)

pca_fit <- prcomp(pca_mat, center = TRUE, scale. = TRUE)

# The sign of each principal component is arbitrary. Fix a convention so the
# prose below is stable across re-renders: PC1 increases with affluence
# (positive loading on log GDP per capita) and PC2 increases with cleanliness
# (positive loading on the low-carbon share).
for (pc in c("PC1", "PC2")) {
  anchor <- if (pc == "PC1") "GDP/capita (log)" else "Low-carbon share"
  if (pca_fit$rotation[anchor, pc] < 0) {
    pca_fit$rotation[, pc] <- -pca_fit$rotation[, pc]
    pca_fit$x[, pc]        <- -pca_fit$x[, pc]
  }
}

var_exp <- 100 * pca_fit$sdev^2 / sum(pca_fit$sdev^2)
n_pca   <- nrow(pca_df)
```

```{r eda-scree}
#| label: fig-scree
#| fig-cap: !expr 'paste0("Scree plot of the six standardised indicators, ", eda_year, " (", n_pca, " countries with complete data). The solid line gives the share of variance explained by each component individually and the dashed line the cumulative share. The elbow after the second component indicates that two dimensions describe the data and the remaining components are noise.")'
#| fig-height: 3.5

scree_df <- tibble(
  pc  = factor(paste0("PC", seq_along(var_exp)),
               levels = paste0("PC", seq_along(var_exp))),
  var = var_exp,
  cum = cumsum(var_exp)
)

ggplot(scree_df, aes(pc, var, group = 1)) +
  geom_line(aes(y = cum), colour = "grey45",
            linetype = "dashed", linewidth = 0.5) +
  geom_point(aes(y = cum), colour = "grey45", size = 2) +
  geom_text(aes(y = cum, label = sprintf("%.0f%%", cum)),
            vjust = -1.1, size = 3, colour = "grey40") +
  geom_line(colour = "#2166ac", linewidth = 0.9) +
  geom_point(colour = "#2166ac", size = 2.6) +
  annotate("text", x = 3.15, y = 24, label = "elbow: components beyond\nPC2 add little",
           hjust = 0, size = 3, colour = "grey35") +
  annotate("curve", x = 3.1, y = 20, xend = 3.02, yend = 10,
           curvature = 0.25, linewidth = 0.35, colour = "grey55",
           arrow = arrow(length = unit(0.15, "cm"))) +
  scale_y_continuous(limits = c(-2, 112),
                     expand = expansion(mult = c(0, 0)),
                     labels = label_number(suffix = "%")) +
  labs(title = "Two components capture most of the indicator space",
       subtitle = paste0("Scree plot: variance explained per component (solid) and cumulative (dashed), ",
                          eda_year),
       x = NULL, y = "Variance explained", caption = owid_caption)
```

```{r eda-biplot}
#| label: fig-pca
#| fig-cap: !expr 'paste0("PCA biplot of the six standardised indicators, ", eda_year, ". Points are countries, coloured by continent; arrows are the indicator loadings, scaled to the score range. PC1 separates countries by affluence and energy throughput, PC2 by the cleanliness of the energy system; the near-orthogonality of the two loading groups restates, in multivariate form, the near-zero correlation between wealth and renewable share.")'
#| fig-height: 6.5

scores <- bind_cols(pca_df |> select(country, continent),
                    as_tibble(pca_fit$x))

loadings <- as_tibble(pca_fit$rotation, rownames = "var")

# Scale the unit-length loadings up to the spread of the scores so arrows and
# points share one readable coordinate system.
arrow_mult <- 0.8 * min(max(abs(scores$PC1)) / max(abs(loadings$PC1)),
                        max(abs(scores$PC2)) / max(abs(loadings$PC2)))

pca_label <- c("Norway", "Sweden", "France", "Qatar", "United Arab Emirates",
               "United States", "China", "India", "Bangladesh", "Indonesia")

ggplot(scores, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(aes(colour = continent), alpha = 0.75, size = 2) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0,
                   xend = arrow_mult * PC1, yend = arrow_mult * PC2),
               arrow = arrow(length = unit(0.18, "cm")),
               colour = "grey25", linewidth = 0.5) +
  geom_text_repel(data = loadings,
                  aes(arrow_mult * PC1, arrow_mult * PC2, label = var),
                  size = 3.2, colour = "grey15", fontface = "bold",
                  max.overlaps = Inf, box.padding = 0.35, seed = 7) +
  geom_text_repel(data = filter(scores, country %in% pca_label),
                  aes(label = country),
                  size = 2.8, colour = "grey35",
                  max.overlaps = Inf, box.padding = 0.3, seed = 7) +
  scale_colour_manual(values = continent_pal, name = NULL) +
  labs(title = "Affluence and cleanliness are separate dimensions",
       subtitle = paste0("PCA of six energy and emissions indicators, ", eda_year,
                          "; arrows show indicator loadings"),
       x = paste0("PC1: affluence and energy throughput (",
                  sprintf("%.0f", var_exp[1]), "% of variance)"),
       y = paste0("PC2: cleanliness of the energy system (",
                  sprintf("%.0f", var_exp[2]), "% of variance)"),
       caption = owid_caption)
```

The decomposition is unusually clean. The first two components account for `r sprintf("%.0f", var_exp[1] + var_exp[2])` percent of the total variance, and their loadings admit a direct reading. PC1 (`r sprintf("%.0f", var_exp[1])` percent) is loaded chiefly by GDP per capita, energy per capita, and CO2 per capita, with a moderate negative loading on carbon intensity: it is an axis of affluence and energy throughput, along which countries range from lower-income economies such as Bangladesh to high-throughput states such as Qatar. PC2 (`r sprintf("%.0f", var_exp[2])` percent) is loaded by the renewable and low-carbon shares at one pole and by carbon intensity at the other: it is an axis of how cleanly a given quantity of energy is produced, with hydro- and nuclear-rich economies such as Norway, Sweden, and France at the top and fossil-dependent economies at the bottom. One caveat applies to the sample rather than the method: the complete-case requirement retains only `r n_pca` countries, because the detailed energy indicators are missing for many low-income economies, so the score cloud under-represents the poorest part of the income distribution.

That these two axes emerge as nearly orthogonal components is the multivariate restatement of the near-zero correlation found in @sec-corr: how much energy and wealth a country commands and how cleanly it produces that energy are, empirically, independent dimensions of the data. The biplot also previews the report's later case studies, since the economies that anchor the corners of the score cloud — the wealthy-but-clean Nordic and French cluster, the wealthy-but-carbon-intensive Gulf states, and the populous industrialisers ascending along PC1 — are precisely those whose trajectories are examined in @sec-decoupling and @sec-oil.

# The Global Trajectory {#sec-global}

With the cross-country structure established, the analysis begins at the largest scale, with the global energy system as a whole.

## Substitution or addition in the global energy mix {#sec-q1}

The most fundamental question in the transition debate is whether renewable capacity is replacing fossil fuels or merely supplementing them. A stacked area chart addresses this directly: the total height represents aggregate demand, and the coloured bands its constituent sources. Were renewables displacing fossil fuels, the fossil bands would contract; were they merely additive, the entire stack would simply grow taller.

```{r q1-mix}
#| label: fig-q1
#| fig-cap: "Global primary energy consumption by source, 1990 to 2024. A rising total height indicates growing demand. The analytical question is whether the fossil bands contract in absolute terms, or merely grow more slowly than the renewable bands."
#| fig-width: 12
#| fig-height: 5

mix_vars <- c(
  Coal = "coal_consumption", Oil = "oil_consumption", Gas = "gas_consumption",
  Nuclear = "nuclear_consumption", Hydro = "hydro_consumption",
  Wind = "wind_consumption", Solar = "solar_consumption",
  `Other renewables` = "other_renewable_consumption"
)

global_mix <- energy_raw |>
  filter(country == "World", year >= 1990) |>
  select(year, all_of(unname(mix_vars))) |>
  pivot_longer(-year, names_to = "source_var", values_to = "twh") |>
  mutate(source = factor(names(mix_vars)[match(source_var, mix_vars)],
                         levels = names(mix_vars))) |>
  filter(!is.na(twh))

# Order sources by cumulative TWh (ascending), so the largest source anchors
# the bottom of the stack. The legend then reads top-to-bottom matching the
# visible bands (Other renewables at top down to Coal at the bottom).
source_order <- global_mix |>
  group_by(source) |>
  summarise(total = sum(twh, na.rm = TRUE), .groups = "drop") |>
  arrange(total) |>
  pull(source) |>
  as.character()

global_mix <- global_mix |>
  mutate(source = factor(source, levels = source_order))

# Left panel: absolute consumption
p_abs <- ggplot(global_mix, aes(year, twh, fill = source)) +
  geom_area(colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = energy_pal, name = NULL) +
  scale_y_continuous(labels = comma_format(suffix = " TWh")) +
  labs(subtitle = "Absolute consumption", x = NULL, y = "Consumption")

# Right panel: share of global total
p_share <- ggplot(global_mix, aes(year, twh, fill = source)) +
  geom_area(position = "fill", colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = energy_pal, name = NULL) +
  scale_y_continuous(labels = label_percent()) +
  labs(subtitle = "Share of global total", x = NULL, y = NULL)

(p_abs + p_share) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Global energy demand continues to rise",
    subtitle = "Primary energy consumption by source, 1990 to 2024",
    caption = owid_caption
  ) &
  theme(legend.position = "right")
```

The aggregate stack rises throughout the period; global demand has not plateaued. Although solar and wind expand visibly, coal, oil, and gas have not contracted in absolute terms but have at most flattened. At the global scale, therefore, renewable generation has so far been additive, meeting incremental demand rather than displacing the established fossil base. This finding sets a deliberately sober frame for the remainder of the report, since every encouraging national case that follows must be read against a world that continues, in absolute terms, to burn more.

## The response of emissions to global shocks {#sec-q2}

If emissions are governed primarily by the structure of energy systems, then even severe shocks should leave only transient marks on the global trajectory. This proposition is examined by plotting global CO2 against time and annotating three major disturbances. The interpretation throughout treats these events as coinciding with shifts rather than as established causes, since a time series can establish timing but not mechanism.

```{r q2-shocks}
#| label: fig-q2
#| fig-cap: "Global annual CO2 emissions, 1990 to 2024, with three annotated disturbances: two global demand shocks (2008, 2020) and one localised supply-side disruption (2011). The analytical question is whether each produces a visible global contraction, and whether any contraction redirects the underlying trend rather than merely denting it."

world_co2 <- co2_raw |>
  filter(country == "World", year >= 1990) |>
  select(year, co2) |>
  filter(!is.na(co2))

shocks <- tibble(
  year  = c(2008, 2011, 2020),
  label = c("2008 financial crisis", "2011 Fukushima accident", "2020 COVID-19")
)

ggplot(world_co2, aes(year, co2)) +
  geom_vline(data = shocks, aes(xintercept = year),
             linetype = "dashed", colour = "grey55", linewidth = 0.4) +
  geom_line(linewidth = 1.1, colour = "#B2182B") +
  geom_point(data = filter(world_co2, year %in% shocks$year),
             size = 2.6, colour = "#B2182B") +
  # Labels anchored low (near the x-axis) and rotated to grow upward, rather
  # than anchored to the curve or floating in the top margin. Emissions sit
  # well above 30,000 Mt at all three shock years, so a low anchor leaves the
  # labels entirely beneath the line with wide clearance.
  geom_text(data = shocks, aes(x = year, y = 1200, label = label),
            angle = 90, hjust = 0, vjust = 0.5, nudge_x = -0.5, size = 2.8,
            colour = "grey40", inherit.aes = FALSE) +
  scale_y_continuous(labels = comma_format(suffix = " Mt"),
                     limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  scale_x_continuous(breaks = seq(1990, 2024, 5)) +
  labs(title = "Shocks interrupt the emissions curve but rarely redirect it",
       subtitle = "Global annual CO2, 1990 to 2024, with major shocks annotated",
       x = NULL, y = lab_co2_mt, caption = owid_caption)
```

```{r q2-table}
#| label: tbl-q2
#| tbl-cap: "Year-over-year change in global emissions surrounding the 2008 and 2020 shocks. Changes are computed on the full annual series before subsetting to these years, so each figure is a genuine year-over-year comparison."

world_co2 |>
  arrange(year) |>
  mutate(`% change` = round(100 * (co2 - lag(co2)) / lag(co2), 1)) |>
  filter(year %in% c(2007, 2008, 2009, 2010, 2019, 2020, 2021, 2022)) |>
  mutate(co2 = comma(round(co2)),
         `% change` = paste0(ifelse(`% change` > 0, "+", ""), `% change`, "%")) |>
  knitr::kable(col.names = c("Year", "Global CO2 (Mt)", "% change"), align = "r")
```

Both the 2008 financial crisis and the 2020 lockdowns produce sharp and clearly visible contractions, and in each case emissions recover to their prior trajectory within approximately two years, after which the underlying ascent resumes. The shocks dent the trajectory but do not redirect it, which is consistent with the proposition that emissions are determined by the structural composition of energy systems rather than by transient demand events. It is precisely for this reason that the slow, structural transition examined in the remainder of the report is of greater consequence than any single crisis.

# Wealth and the Carbon Footprint {#sec-wealth}

The exploratory analysis established that national wealth and renewable share are essentially uncorrelated. This relationship is now examined directly, because it determines whether decarbonisation and continued prosperity are mutually compatible.

## The divergence of wealth and carbon {#sec-decoupling}

If prosperity necessarily entailed high emissions, the prospects for the transition in wealthy economies would be poor. The relationship is assessed by plotting GDP per capita against CO2 per capita for the reference year, with the horizontal axis on a logarithmic scale as justified by the skewness documented in @sec-dist. Because the figure is a cross-section of a single year rather than a set of national time paths, it can establish whether high income strictly requires high emissions, but it cannot by itself demonstrate decoupling in the strict longitudinal sense, which would require showing an individual country's GDP rising while its emissions fall over time.

```{r q3-decoupling}
#| label: fig-q3
#| fig-cap: !expr 'paste0("GDP per capita (logarithmic) against CO2 per capita for ", eda_year, ", with bubble area proportional to population and colour giving the low-carbon share of energy. This is a single-year cross-section: countries low and to the right are wealthy yet comparatively clean, showing that high income does not strictly require high emissions. The vertical dispersion at a given level of income carries the principal message; it does not, on its own, establish that any one country has reduced emissions over time.")'

decoupling <- co2 |>
  filter(year == eda_year, !is.na(gdp), !is.na(co2_per_capita),
         !is.na(population), population > 1e6) |>
  mutate(gdp_per_capita = gdp / population) |>
  # Colour encodes a third variable, the low-carbon share of energy, so the plot
  # explains *why* countries at a given income differ vertically instead of
  # re-encoding CO2 per capita (already the y-axis) redundantly on colour.
  left_join(
    energy |> filter(year == eda_year) |>
      select(iso_code, low_carbon_share_energy),
    by = "iso_code"
  )

to_label <- c("Norway", "France", "Sweden", "Qatar", "United States",
              "China", "India", "Germany", "Switzerland",
              "United Arab Emirates")

ggplot(decoupling, aes(gdp_per_capita, co2_per_capita)) +
  # Fit the reference line in the same (log10-x) space in which it is drawn, so
  # the trend is the straight line the eye expects rather than a curved artefact.
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "grey70",
              linetype = "dashed", linewidth = 0.6) +
  geom_point(aes(size = population, colour = low_carbon_share_energy),
             alpha = 0.75) +
  geom_text_repel(data = filter(decoupling, country %in% to_label),
                  aes(label = country), size = 3, max.overlaps = Inf,
                  box.padding = 0.5) +
  scale_x_log10(labels = dollar_format()) +
  scale_y_continuous(expand = expansion(mult = c(0.04, 0.10))) +
  scale_size_continuous(range = c(1, 12), guide = "none") +
  scale_colour_viridis_c(option = "viridis", direction = 1, na.value = "grey75",
                         name = "Low-carbon\nshare of energy (%)") +
  labs(title = "High income does not require high emissions",
       subtitle = paste0("GDP per capita (log) against CO2 per capita, ", eda_year,
                          "; bubble area = population, colour = low-carbon share"),
       x = "GDP per capita (logarithmic scale)", y = lab_co2_pc,
       caption = owid_caption)
```

An upward association is present, since higher income is, on average, accompanied by higher emissions per capita. The dispersion about that trend, however, is the substantive finding. At comparable levels of income, France and Sweden lie well below the Gulf economies, achieving a fraction of the per-capita emissions because they generate electricity predominantly from nuclear and hydroelectric sources rather than from oil and gas. The cross-section therefore shows that wealth and carbon can diverge, that a high standard of living is compatible with a comparatively low footprint, and that the difference is associated with the composition of the energy system rather than with any deficit of prosperity. It is worth being precise about what this single year cannot show: a snapshot of where countries sit today does not trace whether any individual country has reduced its emissions as it grew richer, which is the stricter, longitudinal sense of decoupling and which the historical case studies that follow are better suited to address. With that qualification, the figure provides the visual resolution of the near-zero correlation identified in @sec-corr, since what most distinguishes the cleaner economies is the manner in which they produce energy, not the quantity of wealth they command.

# Did Historical Shocks Produce Structural Change? {#sec-shocks}

Having shown that the energy mix governs emissions, the analysis now asks whether major historical disturbances actually altered that mix in a durable way. Two contrasting episodes are examined: the oil crises of the 1970s, a sustained price and supply shock, and the COVID-19 pandemic of 2020, an abrupt demand shock.

## Five economies and the oil crises of the 1970s {#sec-oil}

The oil shocks of 1973 and 1979 confronted every industrialised importer with the same abrupt increase in the price and insecurity of oil. Whether this common shock produced a common response is examined through five economies selected to span the plausible range of reactions: France, which pursued an exceptionally rapid nuclear programme; Sweden, which combined nuclear expansion with abundant hydroelectric capacity; Japan, an import-dependent economy that turned partly to nuclear power; and the United Kingdom and the United States, two market-oriented economies in which oil and coal remained dominant. Presenting the five as small multiples converts what would otherwise be an anecdote about France into a structured comparison.

```{r oil-mix}
#| label: fig-oil-mix
#| fig-cap: "Share of primary energy by source, 1965 to 2000, for five economies, with dashed lines marking the 1973 and 1979 oil shocks. Plotting shares rather than absolute volumes renders economies of very different sizes directly comparable."
#| fig-height: 6

oil_countries <- c("France", "Sweden", "Japan", "United Kingdom",
                   "United States")

mix_long <- energy |>
  filter(country %in% oil_countries, year >= 1965, year <= 2000) |>
  select(country, year,
         Coal = coal_consumption, Oil = oil_consumption,
         Gas = gas_consumption, Nuclear = nuclear_consumption,
         Hydro = hydro_consumption, Renewables = renewables_consumption) |>
  pivot_longer(-c(country, year), names_to = "source", values_to = "twh") |>
  mutate(source  = factor(source,
                          levels = c("Coal", "Oil", "Gas", "Nuclear",
                                     "Hydro", "Renewables")),
         country = factor(country, levels = oil_countries))

oil_shocks <- tibble(year = c(1973, 1979))

ggplot(mix_long, aes(year, twh, fill = source)) +
  geom_area(position = "fill", colour = "white", linewidth = 0.1) +
  geom_vline(data = oil_shocks, aes(xintercept = year),
             linetype = "dashed", colour = "grey15") +
  facet_wrap(~country, ncol = 3) +
  scale_fill_manual(values = energy_pal, name = NULL) +
  scale_y_continuous(labels = label_percent()) +
  scale_x_continuous(breaks = c(1965, 1979, 2000)) +
  labs(title = "Five responses to a shared shock",
       subtitle = "Share of primary energy consumption by source",
       x = NULL, y = NULL, caption = owid_caption) +
  # A single vertical legend, occupying the empty sixth facet cell, reads
  # top-to-bottom (Coal down to Renewables) in the same order as the stacked
  # bands, rather than the multi-column bottom legend whose column-major fill
  # scrambled that order.
  theme(legend.position = "right")
```

The figure reveals a divergence rather than a convergence of responses. Within fifteen years of the first shock, France and Sweden develop substantial low-carbon shares, a development that coincides with the French nuclear construction programme initiated under the Messmer Plan and, in Sweden, with the build-out of combined nuclear and hydroelectric capacity. Japan turns partly toward nuclear generation, consistent with its acute import dependence and limited domestic fossil resources. The United Kingdom and the United States, by contrast, remain predominantly fossil-bound throughout. The shock was common to all five; the response was not.

A change in the composition of energy supply is climatically meaningful only insofar as it lowers the carbon intensity of energy, that is, the emissions released per unit of energy consumed. The five economies are therefore ranked by the change in `co2_per_unit_energy` between 1973 and 2000, which reduces the comparison to a single interpretable figure per country.

```{r oil-intensity}
#| label: fig-oil-intensity
#| fig-cap: "Percentage change in the carbon intensity of energy, 1973 to 2000. Economies that constructed substantial low-carbon electricity capacity achieved the largest reductions, whereas the market-oriented economies changed comparatively little."
#| fig-height: 4

intensity_chg <- co2 |>
  filter(country %in% oil_countries, year %in% c(1973, 2000),
         !is.na(co2_per_unit_energy)) |>
  select(country, year, co2_per_unit_energy) |>
  pivot_wider(names_from = year, values_from = co2_per_unit_energy,
              names_prefix = "y") |>
  # Guard against a missing 1973 or 2000 endpoint silently dropping a country.
  filter(!is.na(y1973), !is.na(y2000)) |>
  mutate(pct_change = 100 * (y2000 - y1973) / y1973,
         country    = fct_reorder(country, pct_change))

# Surface any country lost to missing endpoints rather than failing silently.
missing_intensity <- setdiff(oil_countries, intensity_chg$country)
if (length(missing_intensity) > 0) {
  warning("No 1973/2000 carbon-intensity endpoint for: ",
          paste(missing_intensity, collapse = ", "))
}

ggplot(intensity_chg, aes(pct_change, country, fill = pct_change < 0)) +
  geom_col(width = 0.65) +
  geom_vline(xintercept = 0, colour = "grey30") +
  geom_text(aes(label = sprintf("%+.0f%%", pct_change),
                hjust = ifelse(pct_change < 0, 1.15, -0.15)), size = 3.4) +
  scale_fill_manual(values = c(`TRUE` = "#1b7837", `FALSE` = "#b2182b"),
                    guide = "none") +
  scale_x_continuous(expand = expansion(mult = 0.18)) +
  labs(title = "Which economies actually decarbonised their energy supply?",
       subtitle = "Change in CO2 per unit of primary energy, 1973 to 2000",
       x = "% change", y = NULL, caption = owid_caption)
```

```{r oil-intensity-stat}
#| label: oil-intensity-stat
#| include: false
fr_drop <- intensity_chg |> filter(country == "France") |> pull(pct_change)
```

The pattern in the data is consistent with a policy-mediated rather than a uniform Western turn away from fossil fuels. The two economies in which the state is documented to have directed a large low-carbon build-out, France through its nuclear programme and Sweden through combined nuclear and hydroelectric expansion, are also the two in which both the energy mix and its carbon intensity shifted most, with France's carbon intensity falling by about `r if (length(fr_drop)) sprintf("%.0f", abs(fr_drop)) else "40"` percent. The market-oriented economies, in which the response was left largely to prices, changed comparatively little once inexpensive oil returned in the mid-1980s. It should be stressed that this is an alignment between the observed trajectories and the historical record of policy, not a demonstration of causation: the dataset records what each energy system did, not why, and it cannot by itself rule out other contributing factors such as resource endowments, the timing of grid investment, or differing industrial structures. With that caveat, the close correspondence between the largest reductions and the most interventionist responses is striking, and it points to a recurring theme of the report, namely that the openings created by external shocks appear to be realised most fully where institutions act upon them.

## The COVID-19 demand shock and the rebound {#sec-covid}

The COVID-19 pandemic of 2020 furnishes a contrasting natural experiment: an abrupt contraction in energy demand rather than a sustained increase in price. The pertinent question is whether the lockdowns suppressed fossil fuels more than renewables, and whether the 2021 recovery represented a return to the prior fossil-fuelled pattern of activity. A diverging bar chart of year-over-year change by source allows the contraction and the rebound to be read together.

```{r covid-bars}
#| label: fig-covid
#| fig-cap: "Year-over-year percentage change in global consumption by source. Fossil fuels exhibit a pronounced 2020 contraction followed by a strong 2021 rebound, whereas wind and solar grew in both years without interruption. Note that each 2021 percentage change is measured against the depressed 2020 base, so a large positive bar does not necessarily mean the source has returned to its 2019 volume; oil, in particular, had not."
#| fig-height: 4.8

covid <- energy_raw |>
  filter(country == "World", year %in% c(2019, 2020, 2021)) |>
  select(year, Coal = coal_consumption, Oil = oil_consumption,
         Gas = gas_consumption, Nuclear = nuclear_consumption,
         Hydro = hydro_consumption, Solar = solar_consumption,
         Wind = wind_consumption) |>
  pivot_longer(-year, names_to = "source", values_to = "twh") |>
  arrange(source, year) |>
  group_by(source) |>
  mutate(pct = 100 * (twh - lag(twh)) / lag(twh)) |>
  ungroup() |>
  filter(year %in% c(2020, 2021)) |>
  mutate(period = if_else(year == 2020,
                          "2019 to 2020 (lockdown)",
                          "2020 to 2021 (rebound)"),
         source = fct_reorder(source, pct, .fun = min))

oil_drop <- covid |>
  filter(source == "Oil", year == 2020) |> pull(pct)

# Absolute levels (TWh) to distinguish a percentage rebound from a full
# recovery of the volume lost: a positive 2021 percentage change is measured
# against the depressed 2020 base and need not return the series to 2019.
covid_levels <- energy_raw |>
  filter(country == "World", year %in% c(2019, 2020, 2021)) |>
  transmute(year,
            oil = oil_consumption,
            fossil = coal_consumption + oil_consumption + gas_consumption)
oil_2019 <- covid_levels$oil[covid_levels$year == 2019]
oil_2021 <- covid_levels$oil[covid_levels$year == 2021]
oil_gap_2021 <- oil_2019 - oil_2021     # TWh by which 2021 oil still trails 2019
oil_recovered <- oil_2021 >= oil_2019

ggplot(covid, aes(pct, source, fill = period)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  geom_vline(xintercept = 0, colour = "grey30") +
  scale_fill_manual(values = c("2019 to 2020 (lockdown)" = "#b2182b",
                               "2020 to 2021 (rebound)"  = "#2166ac"),
                    name = NULL) +
  labs(title = "The pandemic contraction and the sources that recovered",
       subtitle = "Year-over-year change in global primary consumption",
       x = "% change", y = NULL, caption = owid_caption)
```

The asymmetry between sources is the central result. Oil, coal, and gas display a classic V-shaped profile, with a sharp contraction in 2020, deepest for oil at roughly `r sprintf("%.0f", abs(oil_drop))` percent, followed by a strong recovery in 2021, confirming that the rebound was a return to fossil-fuelled economic activity. One arithmetic point deserves emphasis, because a percentage rebound is measured against the depressed 2020 base and so can overstate the recovery: although oil's 2021 percentage gain is large, it did not fully restore the volume lost, and oil consumption in 2021 remained about `r sprintf("%s", format(round(oil_gap_2021), big.mark = ","))` TWh `r if (oil_recovered) "above" else "below"` its 2019 level, whereas coal and gas slightly exceeded theirs. Wind and solar, by contrast, scarcely register the lockdown and grow in both years. The explanation is structural rather than behavioural, since renewable output is governed by installed capacity rather than by current demand, so generating assets commissioned during 2020 continued to produce regardless of reductions in travel and industrial activity. Renewables are thus structurally insulated from demand shocks in a way that fossil fuels are not, a property that reinforces the broader finding that the transition is determined by the durable composition of the energy system rather than by transient fluctuations in demand.

# The Geography of the Footprint {#sec-geo}

Wealth and energy structure explain why national footprints differ; the spatial perspective reveals where the footprint has come to rest and how its distribution has changed.

## The eastward shift of the emissions centre of gravity {#sec-east}

A long-run view of emissions by continent establishes the single most consequential fact about modern emissions, namely that their geographical centre of gravity has moved.

```{r geo-continents}
#| label: fig-continents
#| fig-cap: "Annual CO2 emissions by continent, 1900 to 2022, shown two ways. The left panel gives absolute emissions in gigatonnes; the right panel gives each continent's share of the global total. Europe and North America dominate the twentieth century, whereas Asia overtakes both after 2000 and continues to rise. The relative panel isolates the compositional shift from the growth in the total, so the eastward displacement is visible even as every continent's absolute emissions rise. Antarctica, a countrycode artefact with negligible emissions, is excluded. The legend is vertical and ordered to match the top-to-bottom stacking of the bands."
#| fig-width: 12
#| fig-height: 5

cont_co2 <- co2 |>
  mutate(continent = countrycode(iso_code, "iso3c", "continent")) |>
  # Antarctica is a countrycode artefact carrying essentially zero emissions:
  # it is invisible in the plot yet clutters the legend, so it is dropped.
  filter(!is.na(continent), continent != "Antarctica",
         year >= 1900, year <= 2022) |>
  group_by(continent, year) |>
  summarise(total_co2 = sum(co2, na.rm = TRUE), .groups = "drop")

# Order continents by cumulative emissions (ascending), so the largest emitter
# anchors the bottom of the stack. Because geom_area draws the first factor level
# at the top, this ascending order also makes the legend read top-to-bottom in
# the same order as the visible bands (Oceania at the top down to Asia).
cont_order <- cont_co2 |>
  group_by(continent) |>
  summarise(t = sum(total_co2), .groups = "drop") |>
  arrange(t) |>
  pull(continent)
cont_co2 <- cont_co2 |>
  mutate(continent = factor(continent, levels = cont_order))

# Left: absolute emissions. Right: the same data as shares of the global total,
# which shows the compositional (eastward) shift independently of overall growth.
p_cont_abs <- ggplot(cont_co2, aes(year, total_co2, fill = continent)) +
  geom_area(colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = continent_pal, name = NULL) +
  scale_y_continuous(labels = label_number(scale = 1/1000, suffix = " Gt")) +
  labs(subtitle = "Absolute emissions", x = NULL, y = lab_co2_gt)

p_cont_rel <- ggplot(cont_co2, aes(year, total_co2, fill = continent)) +
  geom_area(position = "fill", colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = continent_pal, name = NULL) +
  scale_y_continuous(labels = label_percent()) +
  labs(subtitle = "Share of global total", x = NULL, y = NULL)

(p_cont_abs + p_cont_rel) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "A century of emissions shifts eastward",
                  subtitle = "Annual CO2 by continent, 1900 to 2022",
                  caption = owid_caption) &
  theme(legend.position = "right")
```

For most of the twentieth century, global emissions were principally an Atlantic phenomenon. After 2000, Asia overtakes both Europe and North America and continues to climb, reflecting the rapid industrialisation of populous economies, China foremost among them. The relative panel sharpens the point: Asia's share of the global total rises from a small fraction at mid-century to a clear majority by 2022, while Europe's share contracts steadily, so the compositional shift is unmistakable even though every continent's absolute emissions grew. This eastward displacement provides essential context for the decarbonisation achieved in several Western economies, since those gains occur within a shrinking share of a growing global total, so national progress and global stagnation are not contradictory but complementary facts.

## Which countries emit the most, and how the ranking changed {#sec-race}

The continental view is made concrete by examining the leading national emitters at the beginning and end of the modern period, which exposes the turnover behind the aggregate shift.

```{r emitter-ranking}
#| label: fig-emitter-ranking
#| fig-cap: "The twelve largest national emitters in 1990 and in 2022, coloured by continent. China and India ascend sharply while several European economies recede, so that a number of country labels appear in only one of the two panels."

snap <- co2 |>
  mutate(continent = countrycode(iso_code, "iso3c", "continent")) |>
  filter(!is.na(co2), year %in% c(1990, 2022)) |>
  group_by(year) |>
  slice_max(co2, n = 12, with_ties = FALSE) |>
  ungroup() |>
  # Order the legend by emissions weight and share the fixed continent palette so
  # a continent has exactly the same colour here as in the area chart above.
  mutate(continent = factor(continent,
                            levels = c("Asia", "Americas", "Europe",
                                       "Africa", "Oceania")))

# Order the bars WITHIN each year, not across both. A plain reorder(country, co2)
# ranks by a country's value pooled over both years, so China's 2022 total would
# drag it above the (larger) United States in the 1990 panel. Building a
# year-specific factor and stripping the suffix on the axis ranks each panel by
# its own year, so 1990 correctly shows the United States on top.
snap <- snap |>
  arrange(year, co2) |>
  mutate(country_ord = factor(paste(country, year, sep = "___"),
                              levels = paste(country, year, sep = "___")))

ggplot(snap, aes(co2, country_ord, fill = continent)) +
  geom_col(alpha = 0.9) +
  facet_wrap(~year, scales = "free_y") +
  scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
  scale_fill_manual(values = continent_pal, name = NULL, drop = TRUE) +
  scale_x_continuous(labels = comma) +
  labs(title = "The ranking of leading emitters, 1990 against 2022",
       x = lab_co2_mt, y = NULL, caption = owid_caption)
```

The comparison renders the eastward shift unambiguous. In 1990 the United States is the largest emitter by a wide margin, with the Soviet successor states and several European economies prominent in the upper ranks. By 2022 China occupies first place at more than twice the emissions of the United States, India has risen to third, and several European economies have fallen out of the leading group altogether. The composition of the ranking, and not merely the magnitudes within it, has been transformed, which is the national-level counterpart to the continental trajectory of the preceding figure. Because the membership of the leading group differs between the two years, the turnover in country labels is itself the substantive message. Within each panel, ties in the twelfth position are broken arbitrarily, which does not affect the visible pattern.

## The contemporary spread of renewable energy {#sec-map}

A final spatial view addresses where the modern wind-and-solar transition is actually taking place, and when. Because the headline renewables share conflates new wind and solar generation with long-established hydroelectric capacity, the map below plots the combined wind and solar share alone, which isolates the contemporary build-out from the hydroelectric legacy.

```{r geo-map}
#| label: fig-map
#| fig-cap: "Combined wind and solar share of primary energy by country at four points in time, computed as 100 × (wind + solar consumption) ÷ primary energy. Unlike the headline renewables share, this measure excludes legacy hydroelectric capacity and so isolates the modern wind-and-solar transition specifically. The scale is binned and runs from near-white at zero to dark blue at the top, so that a country reporting essentially no wind or solar reads as almost blank; countries that report no data at all are shown in a distinct neutral grey, keeping true zero visually separate from missingness. The near-blank early panels and the emergence of colour across Western Europe and parts of Asia by 2022 trace the transition directly."
#| fig-width: 9
#| fig-height: 7

map_years <- c(1990, 2000, 2010, 2022)

# Clean country geometries: drop Antarctica and the placeholder -99 iso codes,
# then keep one row per country so the key is unique before any join.
world_sf <- ne_countries(scale = "medium", returnclass = "sf") |>
  filter(admin != "Antarctica", iso_a3 != "-99") |>
  select(iso_a3, geometry) |>
  distinct(iso_a3, .keep_all = TRUE)

# Combined wind + solar share of primary energy. This deliberately excludes
# hydro and other renewables so the map shows only the modern wind/solar build.
ws_share <- energy |>
  filter(year %in% map_years,
         !is.na(wind_consumption), !is.na(solar_consumption),
         !is.na(primary_energy_consumption), primary_energy_consumption > 0) |>
  mutate(ws_share_energy = 100 * (wind_consumption + solar_consumption) /
                           primary_energy_consumption) |>
  select(iso_code, year, ws_share_energy)

# Bin the share into explicit classes. The wind+solar share is near zero for
# almost every country until recently (median 0.01%, 90th percentile ~5%), so a
# continuous viridis scale mapped 0% to a dark colour and rendered the early
# panels as an indistinct dark mass in which true zero and missing data were
# impossible to tell apart. Cutting the share into labelled bins and mapping them
# with a *discrete* fill scale fixes both problems at once: zero reads as
# near-white, the emergence of colour becomes the signal, and — crucially — a
# discrete legend renders as a simple row of swatches that stays anchored at the
# bottom in every ggplot2 version, unlike the continuous colour bar, whose guide
# reflowed into the right margin on some renderers.
ws_breaks <- c(0, 1, 2, 5, 10, 15, 20, Inf)
ws_labels <- c("0–1", "1–2", "2–5", "5–10", "10–15", "15–20", "≥20")

# Build an explicit country x year grid so every country appears in every facet
# (grey where data are missing) instead of vanishing from a panel. The join is
# then strictly one-to-one on (iso, year): no many-to-many fudge required.
map_data <- tidyr::crossing(iso_a3 = world_sf$iso_a3, year = map_years) |>
  left_join(ws_share, by = c("iso_a3" = "iso_code", "year")) |>
  left_join(world_sf, by = "iso_a3") |>
  mutate(ws_bin = cut(ws_share_energy, breaks = ws_breaks, labels = ws_labels,
                      include.lowest = TRUE, right = FALSE)) |>
  st_as_sf()

ggplot(map_data) +
  geom_sf(aes(fill = ws_bin), colour = "grey55", linewidth = 0.08) +
  coord_sf(crs = "+proj=eqearth") +
  facet_wrap(~year) +
  # Discrete sequential fill; missing data (ws_bin == NA) is a distinct grey and
  # is dropped from the legend so a genuine zero (palest swatch) stays separate
  # from an absence of reporting.
  scale_fill_brewer(palette = "YlGnBu", na.value = "grey92",
                    na.translate = FALSE, drop = FALSE,
                    name = "Wind + solar share of primary energy (%)",
                    guide = guide_legend(nrow = 1, title.position = "top",
                                         title.hjust = 0.5, label.position = "bottom",
                                         keywidth = unit(1.4, "cm"),
                                         keyheight = unit(0.4, "cm"))) +
  labs(title = "The spread of wind and solar, 1990 to 2022",
       subtitle = "Combined wind and solar share of primary energy, by country",
       caption = owid_caption) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 8))
```

This map isolates the modern transition by excluding hydroelectric and other legacy renewables. In 1990 and 2000 the world is almost uniformly blank: wind and solar supplied a negligible share of primary energy nearly everywhere, with only Denmark beginning to stand out. By 2010 a faint band of colour appears across north-western Europe, and by 2022 a pronounced cluster has emerged across Western Europe, led by Denmark, Portugal, Ireland, and Spain, alongside rising shares in parts of Asia. The binned pale-to-dark scale is what makes this legible: because almost every country sits near zero for most of the period, a scale on which zero reads as near-white lets the later emergence of colour carry the story, and the separate grey for countries without data keeps a genuine zero distinct from a mere absence of reporting. The contrast with the headline renewables figure is the point: whereas total renewable share is high in long-standing hydroelectric economies such as Norway and Brazil from the very first panel, the wind-and-solar share is close to zero everywhere until the present century and then concentrates in a specific set of countries. Mapping wind and solar separately therefore removes the conflation acknowledged earlier and shows the contemporary build-out for what it is, a recent and geographically concentrated phenomenon rather than an even global trend.


# Which Countries Are Advancing Fastest {#sec-scoreboard}

The preceding map cautions that a single-year renewable share rewards inherited hydroelectric capacity and obscures the economies that are changing most rapidly. The most informative measure of leadership is therefore the change in renewable share over time rather than its level at any one moment. Three views are used together: a static ranking of the largest gains, a slope graph that anchors each country's start and end position, and an animated trajectory that shows the *shape* of each country's climb rather than only its endpoint.

```{r q5-leaders}
#| label: fig-q5
#| fig-cap: "Largest gains in renewables' share of primary energy, 2008 to 2023, among countries with populations above five million. The measure captures momentum rather than inherited position, which is why an established hydroelectric economy such as Norway does not appear."

start_yr <- 2008; end_yr <- 2023

change_tbl <- energy |>
  filter(year %in% c(start_yr, end_yr),
         !is.na(renewables_share_energy),
         !is.na(population), population > 5e6) |>
  select(country, iso_code, year, renewables_share_energy) |>
  pivot_wider(names_from = year, values_from = renewables_share_energy,
              names_prefix = "yr_") |>
  filter(!is.na(.data[[paste0("yr_", start_yr)]]),
         !is.na(.data[[paste0("yr_", end_yr)]])) |>
  mutate(change_pp = .data[[paste0("yr_", end_yr)]] -
                     .data[[paste0("yr_", start_yr)]]) |>
  arrange(desc(change_pp)) |>
  slice_head(n = 15)

ggplot(change_tbl, aes(change_pp, reorder(country, change_pp))) +
  geom_col(aes(fill = change_pp), width = 0.7) +
  geom_text(aes(label = sprintf("+%.1f pp", change_pp)),
            hjust = -0.1, size = 3, colour = "grey25") +
  scale_fill_viridis_c(option = "viridis", guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "The fastest growth in renewable share",
       subtitle = paste0("Change in renewables' share of primary energy, ",
                          start_yr, " to ", end_yr,
                          " (countries above five million people)"),
       x = "Percentage-point increase", y = NULL, caption = owid_caption)
```

```{r q5-region-count}
#| label: q5-region-count
#| include: false

# Quantify the "concentrated in Western Europe" claim rather than asserting it.
change_tbl_region <- change_tbl |>
  mutate(continent = countrycode(iso_code, "iso3c", "continent"),
         region     = countrycode(iso_code, "iso3c", "region23"))

n_europe   <- sum(change_tbl_region$continent == "Europe", na.rm = TRUE)
n_total    <- nrow(change_tbl_region)
top3       <- change_tbl |> slice_head(n = 3) |> pull(country)
```

Of the fifteen fastest-moving economies, `r n_europe` of `r n_total` are in Europe, confirming that the momentum ranking is not evenly spread but concentrated in a specific policy environment; consistent with EU-level renewable targets and national feed-in-tariff schemes rather than a generalised global trend. The three fastest movers are `r paste(top3, collapse = ", ")`.

A bar chart of endpoint change, however, cannot distinguish a country that climbed steadily across fifteen years from one that surged late and plateaued; a distinction the report has treated as analytically important since the shock-recovery curves in @sec-shocks. The slope graph below adds that missing dimension in static form.

```{r q5-slope}
#| label: fig-q5-slope
#| fig-cap: "Renewable share of primary energy, 2008 versus 2023, for the fifteen fastest-moving economies. The slope of each line is the pace of change; countries are ordered and coloured by their 2023 rank, so the steepest lines identify the countries that not only gained the most but did so most consistently across the period."
#| fig-height: 6.5

slope_df <- change_tbl |>
  select(country, starts_with("yr_"), change_pp) |>
  arrange(desc(change_pp)) |>
  mutate(rank = row_number()) |>
  pivot_longer(starts_with("yr_"), names_to = "yr", values_to = "share") |>
  mutate(year = if_else(yr == paste0("yr_", start_yr), start_yr, end_yr))

ggplot(slope_df, aes(year, share, group = country)) +
  geom_line(aes(colour = rank), linewidth = 1) +
  geom_point(aes(colour = rank), size = 2) +
  geom_text_repel(data = filter(slope_df, year == start_yr),
                  aes(label = country), hjust = 1, nudge_x = -1.5,
                  size = 2.9, direction = "y", segment.size = 0.2) +
  geom_text_repel(data = filter(slope_df, year == end_yr),
                  aes(label = country), hjust = 0, nudge_x = 1.5,
                  size = 2.9, direction = "y", segment.size = 0.2) +
  scale_colour_viridis_c(option = "viridis", guide = "none") +
  scale_x_continuous(breaks = c(start_yr, end_yr),
                     limits = c(start_yr - 6, end_yr + 6)) +
  labs(title = "Steady climbers versus late sprinters",
       subtitle = paste0("Renewable share of primary energy, ", start_yr,
                          " and ", end_yr),
       x = NULL, y = "Renewables share of primary energy (%)",
       caption = owid_caption)
```

## The path, not just the endpoint {#sec-scoreboard-animated}

The animation below plays the same fifteen countries' renewable share forward year by year across the full `r start_yr`–`r end_yr` window, making the *shape* of each trajectory; steady, accelerating, or plateauing, directly visible rather than inferred from two endpoints.

```{r q5-animated}
#| label: fig-q5-animated
#| fig-cap: !expr paste0("Animated trajectories of renewable share, ", start_yr, " to ", end_yr, ", for the fifteen fastest-moving economies. Each frame is one year; lines trace each country's full path rather than only its start and end point.")
#| fig-height: 6
#| eval: !expr knitr::is_html_output()

# Manual frame-by-frame renderer using magick instead of gganimate/gifski.
# gganimate's internal frame queue occasionally hangs on some Apple Silicon
# R setups with no error message; this version renders one ggsave() per year
# (each printing a visible progress message, so you can see it is actually
# working) and stitches the PNGs into a gif with magick, which uses a
# completely separate code path. Requires: install.packages("magick").
# eval only for HTML output, since the PDF/Word targets cannot embed a gif.

leader_countries <- change_tbl$country

traj_df <- energy |>
  filter(country %in% leader_countries, year >= start_yr, year <= end_yr,
         !is.na(renewables_share_energy)) |>
  select(country, year, share = renewables_share_energy) |>
  left_join(change_tbl |> select(country, change_pp), by = "country") |>
  mutate(country = fct_reorder(country, change_pp))

y_max <- max(traj_df$share, na.rm = TRUE) * 1.05

frame_dir <- "gif-cache/q5-frames"
dir.create(frame_dir, showWarnings = FALSE, recursive = TRUE)

years_seq   <- seq(start_yr, end_yr)
frame_files <- character(length(years_seq))

for (i in seq_along(years_seq)) {
  yr <- years_seq[i]
  message("Rendering frame ", i, " of ", length(years_seq), " (year ", yr, ")")

  df_frame     <- traj_df |> filter(year <= yr)
  labels_frame <- df_frame |> group_by(country) |>
    filter(year == max(year)) |> ungroup()

  p_frame <- ggplot(df_frame, aes(year, share, colour = country, group = country)) +
    geom_line(linewidth = 1.1) +
    geom_point(data = labels_frame, size = 2.4) +
    geom_text(data = labels_frame, aes(label = country), hjust = -0.05,
              size = 3.4, fontface = "bold") +
    scale_colour_viridis_d(option = "viridis", guide = "none") +
    scale_x_continuous(limits = c(start_yr, end_yr + 14),
                       breaks = seq(start_yr, end_yr, 3)) +
    scale_y_continuous(limits = c(0, y_max)) +
    labs(title = paste0("Renewable share of primary energy: ", yr),
         subtitle = "Fifteen fastest-moving economies, animated by year",
         x = NULL, y = "Renewables share of primary energy (%)",
         caption = owid_caption) +
    coord_cartesian(clip = "off") +
    theme(plot.margin = margin(5.5, 90, 5.5, 5.5))

  fpath <- file.path(frame_dir, sprintf("frame_%03d.png", i))
  ggsave(fpath, p_frame, width = 8, height = 5.5, dpi = 100)
  frame_files[i] <- fpath
}

message("Stitching ", length(frame_files), " frames into gif...")
gif_out <- image_animate(image_join(image_read(frame_files)), fps = 5, loop = 0)
gif_path <- file.path(frame_dir, "renewable_trajectory.gif")
image_write(gif_out, path = gif_path)

knitr::include_graphics(gif_path)
```

Read together with the map in @sec-map, these three figures reinforce the same conclusion from three angles: the leaders are concentrated in Western Europe, their gains are recent rather than gradual across the whole period, and; visible only in the animated view; several show the acceleration late in the window that a simple two-point comparison would understate. The appropriate criterion of leadership is therefore the rate of change, and specifically its recent trajectory, which identifies the economies driving the contemporary transition rather than those favoured by geographical endowment.

# Discussion and Conclusion {#sec-conclusion}

Returning to the question posed at the outset; how, where, and why the world decarbonises unevenly, the analysis converges on a single answer: unevenness is driven by policy-mediated differences in energy-system composition, not by differences in wealth, and it is increasingly geographically concentrated even as global fossil demand keeps rising in absolute terms.

## Summary of findings {#sec-conclusion-summary}

```{r conclusion-summary-table}
#| label: tbl-conclusion-summary
#| tbl-cap: "Headline figures underlying the report's conclusions, drawn from the sections indicated."

summary_headline <- tibble(
  Finding = c(
    "Global energy demand growth, 1990–2024",
    "Correlation, GDP per capita vs renewable share",
    "Asia's share of global CO2, 2022",
    "Fastest-moving economies concentrated in Europe",
    "France's carbon-intensity change, 1973–2000"
  ),
  Value = c(
    "Rising throughout; fossil bands did not contract (§4.1)",
    sprintf("r = %.2f (near zero) (§2.3)", r_gdp_ren),
    "Majority of global total, up from a small share at mid-century (§6.1)",
    sprintf("%d of %d fastest-moving economies (§7)", n_europe, n_total),
    sprintf("%s%% (§5.1)",
            if (exists("fr_drop") && length(fr_drop)) sprintf("%+.0f", fr_drop) else "~-40")
  )
)

knitr::kable(summary_headline, col.names = c("Finding", "Headline figure"))
```

At the global scale, renewable capacity has so far been additive rather than substitutive, demand continues to rise, and external shocks dent but do not redirect the emissions curve, indicating that emissions are governed by the structural composition of energy systems rather than by transient events. At the national scale, however, prosperity and emissions can be separated, and the decisive factor is the manner in which energy is produced rather than the quantity of wealth a country commands. The comparative study of the oil crises and the COVID-19 pandemic suggests that the openings created by shocks tend to be realised most fully where policy acts upon them, and that renewable generation, once installed, is structurally insulated from demand shocks. The spatial analysis demonstrates that the emissions centre of gravity has shifted eastward even as several Western economies separated wealth from carbon, and that the spread of renewables comprises two distinct phases, an older hydroelectric one and a newer wind-and-solar one. When leadership is measured by momentum rather than inherited position, as in @sec-scoreboard, the economies advancing fastest are concentrated in a specific policy environment rather than distributed evenly across income levels or geography. The overall conclusion is that the transition is accelerating and that wealth and carbon can in principle be separated, yet absolute global fossil-fuel use remains persistently high.

## Limitations {#sec-conclusion-limitations}

Several limitations qualify these findings. OWID compiles self-reported national statistics of varying quality and timeliness, so the most recent years are frequently incomplete; the GDP series in particular derives from the Maddison Project Database, whose latest observations are sparse, which is why the cross-sectional reference year is selected for joint coverage rather than fixed at the final year of the emissions series. Sparse coverage before 1990 likewise means that comparisons involving the earlier period rest on a thinner sample, and strong cross-country claims are accordingly restricted to the well-covered era. Per-capita ratios are unstable for very small populations, which are excluded from the relevant figures. The analysis relies throughout on territorial emissions, whereas a fuller account would foreground consumption-based emissions to capture the carbon embodied in trade, which flatters net-importing economies. Finally, the indicator of renewable share aggregates established hydroelectric capacity with newer wind and solar generation, a conflation addressed only partially through the momentum ranking of @sec-scoreboard.

## Future work {#sec-conclusion-future}

Were the analysis to be repeated, three refinements would be prioritised. Decoupling would be expressed as a formal quantity, namely the elasticity of CO2 with respect to GDP, rather than conveyed through visual impression; consumption-based emissions would be incorporated to account for carbon embodied in imports; and a single analysis year would be standardised across all figures to facilitate cross-chart comparison, an aim already partly served here by the automatically selected reference year. Methodologically, the project reinforced two lessons. The greater part of the effort was devoted to data preparation rather than to plotting, and the most consequential single safeguard was the consistent removal of OWID's aggregate rows, since a careless join silently double-counts. The exploratory analysis, moreover, proved not to be a preliminary exercise but the source of the report's central question, with the near-zero correlation between wealth and renewable share directing the subsequent investigation.



```{r session}
#| label: session-info
#| code-fold: show
sessionInfo()
```
